Solutions to the potential for unintended conservation consequences of MPAs

### 2021-07-22 documents/03_marlin_paper.Rmd
first real cut of experiments
Retool experiments (#4)
* redesigned experiments around simpler structure

* first real cut of experiments

---
title: "Solutions to the potential for unintended conservation consequences of MPAs"
author:
    - Daniel Ovando
    - Darcy Bradley
    - Lennon Thomas
    - Echelle Burns
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2: default
  bookdown::html_document2: default
bibliography: references.bib
params:
  run_name: ["v1.0"]
---

@@

assign(str_remove(rdses[8],".rds"), read_rds(file.path(results_path,rdses[2])))

@@

# a bit of housekeeping to avoid saving a massive object
tmp <- raw_experiment_results  %>%
  map( ~ .x$obj)

@@

total_results <- total_experiment_results

rm(total_experiment_results)

results <- placement_experiment %>%
  mutate(temp = raw_experiment_results) %>%
  group_by_at(colnames(.)[!colnames(.) %in% c("temp", "prop_mpa")]) %>%
  nest() %>%
  ungroup() %>%
  mutate(placement_id = 1:nrow(.)) %>%
  unnest(cols = data) %>%
  mutate(xid = map(raw_experiment_results, names)) %>%
  unnest(cols = c(temp, xid)) %>%
  mutate(obj = map(temp, "obj"),
         xid = as.numeric(xid)) %>%
  unnest(cols = obj)

@@

warning("move placement experiement id back above run_experiments")

# total_experiment <- placement_experiment %>%
#   left_join(experiments, by = 'xid')

# results <- results %>%
#   left_join(placement_experiment)

@@

results <- results %>%
  group_by(placement_id, xid, critter) %>%
  mutate(bau_biodiv = biodiv[prop_mpa == 0],
         bau_econ = econ[prop_mpa == 0])

excols <-
  experiments %>%
  select(xid, data) %>%
  unnest(cols = data) %>%
  select(-habitat,-critter, -scientific_name, -centroid) %>%
  ungroup() %>%
  unique()

@@

@@

However, the dynamic nature of pelagic systems also creates a risk of unintended conservation consequences of MPAs. When an MPA is implemented on former fishing grounds, fishing fleets have two choices: exit the fishery or move to new fishing grounds. When MPAs redistribute effort away from one area and into another, they can accidentally increase fishing effort on species found within the new fishing grounds [@abbott2012]. This problem of policy "leakage" is well known in many environmental fields [@citations] but the general properties of this problem have received relatively little attention in the theoretical or empirical MPA literature [see @abbott2012;@savina2008;fulton2011a;@baum2003 for notable exceptions]. We fill this gap by

@@

@hastings2017 employed a simply model to derive conditions under which
MPAs would benefit bycatch species. However they also pointed out that
even small increases in model complexity quickly make analytical
solutions to the effects of MPAs intractable. As such, we based our numerical
analysis on a spatially explicit, multi-species, multi-fleet,
bio-economic model we call
[`marlin`](https://github.com/DanOvando/marlin). Models of MPAs span a
wide range of complexity, from simple analytical tools to end-to-end
ecosystem models [@fulton2015]. `marlin` follows a "models of
intermediate complexity" (MICE) philosophy [@plagányi2014] targeted
specifically around the role of technical interactions across species.
Many MPA models simulate the impacts of MPAs on a single-species level [@ovando2021XX]. This can be useful for for example examining the ways in which species-specific life history traits interact with fleet dynamics to influence MPA outcomes. However, bycatch problems are driven by technical interactions, in which fishing effort for a target species incidentally captures another species xx how do we want to talk about this in here? Like usually incidentally, but clearly sometimes people try and catch sharks too xx. `marlin` accounts for this by allowing the user to simulate an arbitrary number of fleets capable of catching an arbitrary number of species in a spatial area. Different fleets can place different economic value on different species, while also having species-specific selectivity functions. This allows us to simulate bycatch dynamics of MPAs in pelagic ecosystems.

@@

Given the complexity of the operating model used here, a complete sensitivity analysis across all parameters is not practical. Instead, we focus on the role of XX key attributes of pelagic ecosystems: the degree of spatial overlap between populations, the presence of seasonal migrations, and the potential for ontogenetic shifts in habitat use, and degree of fishing pressure on each species.

## Species & Fleets

@@

### Ontogenetic Shifts

`marlin` allows for different habitats to be used by adults vs. recruits. For scenarios with ontogenetic shifts ,we assign recruits to a habitat grid that is the inverse of $habitat_{p,s}$. In this way organisms enter the model living on one area, and then transition to adult habitat at older life stages.

XX this isn't very interesting at this point since movement is so high so I vote we get rid of it for now XX.

@@

## Experiments: How severe and undere what conditions are unintended conservation consequences? (clearer name for these)

@@

```{r exp-results, fig.cap="Results of simulation experiments on change in SSB/SSB0 (y-axis) as a function of MPA size (x-axis). Color indicates number of simulations in log-10 space that fell into a particular effect size at a given MPA size"}
tmp <- results %>%
  left_join(excols, by = "xid") %>%
  ggplot(aes(prop_mpa, biodiv - bau_biodiv)) +
  geom_hline(aes(yintercept = 0), linetype = 2, color = "red") +
  geom_bin2d() +
  facet_wrap( ~ critter) +
  scale_x_continuous(labels = scales::percent, name = "% MPA") +
  scale_y_continuous(name = "Change in SSB/SSB0") + 
  scale_fill_viridis_c(trans = "log10")

tmp

Moving to expressing things are percentage of species with some / significant loss

key result: Assuming perfect information, degree of habitat matters buts its messy. Both the depletion and target_fishing strategies are capable of completely avoiding losses if all species share the same habitat. As habitat non-overlap (xx better phrase) inceases, this gets harder to do, and both incur some losses. All of the other placement strategies exhhibit some losses, even with perfect overlap, at some MPA size.

```{r hab-effect, fig.cap="Precent of species with some degree of loss in SSB as a function of MPA size, MPA design strategy, and degree of habitat non-overlap (sigma_c), come up with a better name for this. Perfect information, no seasonal movement, all species considered in planning."}
fig_3 + 
  scale_y_continuous(labels = scales::percent, name = "% of Species with Loss in SSB/SSB0") +
  scale_color_discrete(name = "Habitat Non-Overlap") + 
  theme(legend.position = "top")
```

moving away from perfect information though, things get messier: now just looking at the same plot but with all the other options turned on things get messy quickly.

```{r mess-effect, fig.cap="Precent of species with some degree of loss in SSB as a function of MPA size, MPA design strategy, and degree of habitat non-overlap (sigma_c), come up with a better name for this. Rows indicate presence or absence of seasonal movement."}

tmp <- total_results %>%
  ggplot(aes(prop_mpa, loss, color = factor(sigma_centroid))) +
  geom_jitter(alpha = 0.1) +
  geom_smooth() +
  facet_grid(seasonal_movement ~ placement_strategy) + 
    scale_color_discrete(name = "Habitat Non-Overlap") + 
    scale_y_continuous(labels = scales::percent, name = "% of Species with Loss in SSB/SSB0")+
  theme(legend.position = "top")

tmp
```

same thing but now for SSB loss greater than or equal to 10%. Maybe should make all figure this? Basically, pretty common to cause some technical loss, but it's usually pretty marginal and effectively zero, so no benefit but not really much loss either. Significant losses can happen but less common, though throughout need to be really careful about being clear that this nothign about the probability of any one of these states of nature.

```{r signif-mess-effect, fig.cap="Percent of species with more than 10% loss in SSB/SSB0 as a function of MPA size, MPA design strategy, and degree of habitat non-overlap, come up with a better name for this. Rows indicate presence or absence of seasonal movement."}

tmp <- total_results %>%
  ggplot(aes(prop_mpa, signif_loss, color = factor(sigma_centroid))) +
  geom_jitter(alpha = 0.1) +
  geom_smooth() +
  facet_grid(seasonal_movement ~ placement_strategy) + 
    scale_color_discrete(name = "Habitat Non-Overlap") + 
      scale_y_continuous(labels = scales::percent, name = "% of Species with more than 10% Loss in SSB/SSB0")+

  theme(legend.position = "top")

tmp
```

Huh the seasonal migration thing is counterintuitive, but now that I think about it maybe makes sense, since with migrations every species gets some protection and some fishing, rather than some getting a lot of protection and others a lot more fishing.

My prior: that the magnitude of the benefits will be smaller with seasonal migration. So, you get a sort of tradeoff where on the plus side seasonal migrations means less change of unitended consequences, but less overall benefits as well

Yep that mostly checks.

```{r signif-mess-biodiv-effect, fig.cap="Percent of species with more than 10% loss in SSB/SSB0 as a function of MPA size, MPA design strategy, and degree of habitat non-overlap, come up with a better name for this. Rows indicate presence or absence of seasonal movement."}

tmp <- total_results %>%
  ggplot(aes(prop_mpa, delta_biodiv, color = factor(sigma_centroid))) +
  geom_jitter(alpha = 0.1) +
  geom_smooth() +
  facet_grid(seasonal_movement ~ placement_strategy) + 
    scale_color_discrete(name = "Habitat Non-Overlap") + 
      scale_y_continuous(name = "Total change in SSB/SSB0")+

  theme(legend.position = "top")

tmp

The other thing to consider is quality of information... weird results so far.

Poorer information decreases benefits in some cases...

```{r mess-biodiv-voi-effect, fig.cap="Precent of species with some degree of loss in SSB as a function of MPA size, MPA design strategy, degree of observation error. Rows indicate presence or absence of seasonal movement."}

tmp <- total_results %>%
  ggplot(aes(prop_mpa, delta_biodiv, color = factor(placement_error))) +
  geom_jitter(alpha = 0.1) +
  geom_smooth() +
  facet_grid(seasonal_movement ~ placement_strategy) + 
    scale_color_discrete(name = "Placement Error") + 
    scale_y_continuous(name = "Change in SSB/SSB0")+
  theme(legend.position = "top")

tmp
```

But has a confusing effect on unintended consequences. I guess makes some sense if unintended consequences are a consequence of very targeted interventions, and the greater observation error becomes, the close MPA placement gets to random, and the closer it gets to random, the less likely you are to start hammering one species while benefiting another, rather all just get some weaker protection? Kind of interesting if so.

```{r mess-loss-voi-effect, fig.cap="Precent of species with some degree of loss in SSB as a function of MPA size, MPA design strategy, degree of observation error. Rows indicate presence or absence of seasonal movement."}

tmp <- total_results %>%
  ggplot(aes(prop_mpa, loss, color = factor(placement_error))) +
  geom_jitter(alpha = 0.1) +
  geom_smooth() +
  facet_grid(seasonal_movement ~ placement_strategy) + 
    scale_color_discrete(name = "Placement Error") + 
    scale_y_continuous(labels = scales::percent, name = "% with loss in SSB/SSB0")+
  theme(legend.position = "top")

tmp
```

@@

```{r, fig.cap="Percent of species with some conservation loss as a function of MPA size. Points are draws from rule-of-thumb scenarios, lines are results of optimization routine across a gradient of conservation weighting."}
total_experiment_obj %>%
  filter(prop_mpa > 0) %>%
  ggplot(aes(prop_mpa, loss)) +
    geom_point() +
  geom_smooth(data = total_opt_obj, aes(p_protected / 100, loss, color = alpha, group = alpha), se = FALSE) +
  scale_x_continuous(labels = scales::percent, name = "Area in MPA") +
  scale_y_continuous(labels = scales::percent, name = "% Experiments With Conservation Loss") +
  facet_wrap(~placement_strategy) +
  scale_color_viridis_c(name = "Conservation Weighting") + 
  theme(legend.position = "top")
```

@@

cs_fig_11

@@

cs_fig_10 + 
  theme(legend.position = "top")
```

>>>>

---
title: "Solutions to the potential for unintended conservation consequences of MPAs"
author:
    - Daniel Ovando
    - Darcy Bradley
    - Lennon Thomas
    - Echelle Burns
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2: default
  bookdown::html_document2: default
bibliography: references.bib
params:
  run_name: ["v0.5"]
---

@@

@@

@@

results <- placement_experiments %>%
  mutate(temp = raw_experiment_results) %>%
  mutate(state_id = map(raw_experiment_results, names)) %>%
  unnest(cols = c(temp, state_id)) %>%
  mutate(obj = map(temp, "obj"),
         state_id = as.numeric(state_id)) %>%
  unnest(cols = obj) %>%
  mutate(
    critter = fct_recode(
      critter,
      "Shark" = "carcharhinus longimanus",
      "Tuna" = "katsuwonus pelamis",
      "Marlin" = "kajikia audax"
    )
  )

@@

@@

results <- results %>%
  group_by(placement_id, state_id, critter) %>%
  mutate(bau_biodiv = biodiv[prop_mpa == 0],
         bau_econ = econ[prop_mpa == 0])

excols <-
  state_experiments %>%
  select(state_id, data) %>%
  unnest(cols = data) %>%
  select(state_id:sigma_hab) %>%
  ungroup() %>%
  unique()

total_results <- results %>%
  group_by(
    placement_id,
    placement_strategy,
    prop_critters_considered,
    placement_error,
    state_id,
    prop_mpa
  ) %>%
  summarise(
    loss = mean(biodiv < bau_biodiv),
    n_loss = sum(biodiv < bau_biodiv),
    signif_loss = mean(biodiv < (0.9 * bau_biodiv)),
    total_loss = sum(biodiv - bau_biodiv),
    econ = sum(econ),
    biodiv = sum(biodiv),
    bau_econ = sum(bau_econ),
    bau_biodiv = sum(bau_biodiv)
  ) %>%
  ungroup() %>%
  mutate(delta_biodiv = biodiv / bau_biodiv - 1,
         abs_delta_biodiv = biodiv - bau_biodiv,
         delta_econ = econ / bau_econ - 1) %>%
  left_join(excols, by = "state_id")
```

** something to point out: increased catches are a paradox, either come from rebuilding stock or concentrated fishing, but trajectory can tell you something. not possible for MPAs to increase catches without decreasing catches first if they are rebuilding? **

Seasonal movemetn is kind of interesting, basically just dilutes the extremes

less resilient species are both more in need of protection and easier to accidentally fish harder

@@

Empirically, how do you differentiate increase in catch from fishing harder vs increases in catch from rebuilding: rebuilding has to have a decrease at first

@@

However, the dynamic nature of pelagic systems also creates a risk of unintended conservation consequences of MPAs. When an MPA is implemented on former fishing grounds, fishing fleets have two choices: exit the fishery or move to new fishing grounds. When MPAs redistribute effort away from one area and into another, they can accidentally increase fishing effort on species found within the new fishing grounds [@abbott2012]. This problem of policy "leakage" is well known in many environmental fields [@citations] but the general properties of this problem have received relatively little attention in the theoretical or empirical MPA literature [see @abbott2012;@savina2008;fulton2011a;@baum2003 for notable exceptions]. We fill this gap by examining the general range of conditions...

@@

@hastings2017 employed a simple model to derive conditions under which
MPAs would benefit bycatch species. However they also pointed out that
even small increases in model complexity quickly make analytical
solutions to the effects of MPAs intractable. As such, we based our numerical
analysis on a spatially explicit, multi-species, multi-fleet,
bio-economic model we call
[`marlin`](https://github.com/DanOvando/marlin). Models of MPAs span a
wide range of complexity, from simple analytical tools to end-to-end
ecosystem models [@fulton2015]. `marlin` follows a "models of
intermediate complexity" (MICE) philosophy [@plagányi2014] targeted
specifically around the role of technical interactions across species.
Many MPA models simulate the impacts of MPAs on a single-species level [@2021]. This can be useful for for example examining the ways in which species-specific life history traits interact with fleet dynamics to influence MPA outcomes. However, bycatch problems are driven by technical interactions, in which fishing effort for a target species incidentally captures another species xx how do we want to talk about this in here? Like usually incidentally, but clearly sometimes people try and catch sharks too xx. `marlin` accounts for this by allowing the user to simulate an arbitrary number of fleets capable of catching an arbitrary number of species in a spatial area. Different fleets can place different economic value on different species, while also having species-specific selectivity functions. This allows us to simulate bycatch dynamics of MPAs in pelagic ecosystems.

@@

Given the complexity of the operating model used here, a complete sensitivity analysis across all parameters is not practical. Instead, we focus on the role of life history, fishing pressure,

### Species & Fleets

@@

@@

We focused first on the range of impacts that MPAs can have on mature biomass across the range of variables explored. To illustrate, we pulled out two example simulations, one where MPAs provide relatively large benefits across all three species, and a second where MPAs actually produced substantial losses in some species....

```{r fig-info, fig.cap="Illustration of ways in which variables can interact to affect MPA outcomes."}

Looking at the full `r n_distinct(total_results$placement_id) * n_distinct(total_results$state_id)` experiments run, the results were highly variable, with MPAs capable of producing substantial conservation gains to substantial conservation losses and everything in between. We do not focus into the individual effects of particular variables as we found them to be highly dependent on the exact question asked and the set of data included, and since we cannot ascribe any actual likelihood to any individual state of nature simulated here.

However, three important features do emerge. The first is that life history plays an important role in the potential for unintended conservation consequences from MPAs. All else equal, MPAs were much less likely to produce losses in spawning biomass for our tuna-like species than the marlin or shark-like species. Given that within the broad life history traits we do vary both adult and larval movement, as well as the timing of density dependence, we believe this result is mostly due to the high resilience (high steepness, low age at maturity, fast growth) of the tuna-like species relative to the marlin and shark-like species.

The second is that design matters...

The third is that extreme outcomes (both positive and negative), particularly at lower MPA sizes, are largely driven by the presence of small range sizes interacting with seasonal migrations. Across species simulations without seasonal migrations and small ranges were much more likely to produce extreme results than those with seasonal migrations and/or larger ranges. Presence of seasonal migrations acts as a dilutant: As a given static MPA only provides protection, or displaces fishing effort, onto a particular area in only half of the year, both the benefits and costs of MPAs are smaller. When seasonal migrations are not present and the species range is small, MPAs are more capable of providing either large benefits or large losses.

@@

@@

```{r, fig.cap="Percent of species with some conservation loss as a function of MPA size. Points are draws from rule-of-thumb scenarios, lines are results of optimization routine across a gradient of conservation weighting."}

@@

@@
